---
  title: "Background: EFA models using lavaan"
  author: "Erik-Jan van Kesteren & Rogier Kievit"
  geometry: margin=4cm
  output:
    rmarkdown::html_document:
      toc: false
      theme: flatly
      highlight: zenburn
---

[Go back](../efast)

```{r efast_true, include=FALSE}
library(efast)
library(corrplot)
library(GPArotation)
options(width = 70)
set.seed(45)
```

The `lavaan`[^lavaan] syntax since version `0.6.4` allows for exploratory blocks for latent variables. The EFAST package builds on this functionality to combine exploratory latent variable models (such as EFA) with structural parameters (in EFAST the residual covariance structure). Here is how to create a basic 3-factor EFA model in `lavaan` using the `x1 - x9` variables from the famous Holzinger and Swineford data included in the `lavaan` package:

```{r efa}
data("HolzingerSwineford1939")
efa_model <- "
  efa('block1')*F1 =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
  efa('block1')*F2 =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
  efa('block1')*F3 =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
"
```

The `efa('block1')` modifier indicates that the factors `F1`, `F2`, and `F3` comprise an exploratory block (with the name `block1`) and thus that we do not exactly know the loading structure of the variables `x1 - x9` on these three factors.

The EFA model can then be estimated as follows, (releasing the residual variances and implying the EFA constraints):

```{r efa2}
efa_hs39 <- lavaan(
  model          = efa_model, 
  data           = HolzingerSwineford1939,  
  auto.var       = TRUE, 
  auto.efa       = TRUE
)
```

In the results, among the parameter estimates we will see the pattern of factor loadings resulting from the oblique geomin rotation:

```{r efa3}
summary(efa_hs39)
```
In this summary, we can see that the geomin rotation has produced the structure we expect for this data: the first factor is most associated with the first three variables, the second with `x4 - x6`, and the last three variables load mostly on the third factor.

This is also the same structure we get when we run an oblique geomin-rotated exploratory factor analysis using the same data in the `psych` package (except for the ordering of the factors):

```{r efapsych}
efa_psych <- psych::fa(r = HolzingerSwineford1939[, 7:15], nfactors = 3, rotate = "geominQ")
loadings(efa_psych)
```

The EFAST package implements this exact pattern, and allows for extracting the loadings similar to the `psych` package:

```{r efaefast}
efa_efast <- efast_efa(data = HolzingerSwineford1939[, 7:15], M = 3)
efast_loadings(efa_efast)
``` 

Bonus: a shorter way to write the same syntax would be:

```
efa('block1')*F1 + 
efa('block1')*F2 + 
efa('block1')*F3 =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
```

which is how the EFAST package works (see [Extending EFAST models](./extend_efast#full-syntax)).


[^lavaan]: Yves Rosseel (2012). lavaan: An R Package for Structural Equation Modeling. Journal of
  Statistical Software, 48(2), 1-36. URL http://www.jstatsoft.org/v48/i02/.
